---
title: "Historical Trends in R Package Structure and Interdependency on CRAN"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: flatly
bibliography: pkgstats.bib
header-includes: 
    - \usepackage{tikz}
    - \usetikzlibrary{arrows}
vignette: >
  %\VignetteIndexEntry{pkgstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r pkg-load, echo = FALSE, message = FALSE}
library (pkgstatsAnalyses)
library (ggplot2)
library (patchwork)
library (dplyr)
library (tidyr)
library (igraph)
here <- here::here ()
```

```{r dl-data, echo = FALSE}
path <- file.path (here, "data-raw")
if (!dir.exists (path)) {
    dir.create (path, recursive = TRUE)
}

u <- paste0 ("https://github.com/mpadge/pkgstats-analyses/",
             "releases/download/v0.0.1/")

f_r <- file.path (path, "pkgstats-results.Rds")
if (!file.exists (f_r)) {
    u_r <- paste0 (u, "pkgstats-results.Rds")
    download.file (u_r, f_r)
}

f_p <- file.path (path, "pypi.Rds")
if (!file.exists (f_p)) {
    u_p <- paste0 (u, "pypi.Rds")
    download.file (u_p, f_p)
}
```


# Introduction

A key component of the success of Open Source Software (OSS) is package
managers [@pkg-mngrs-2018], which effectively serve as curated repositories of
independently-contributed software. Package managers exist both for specific
operating systems [@pkg-mngrs-2018] and specific computer languages such as
python [@new-pypi-2018], LaTeX [@ctan-wiki], JavaScript [@npm-2020], and many
others. This work analyses all current and historically archived packages in
the CRAN (Comprehensive R Archive Network) package manager for the R language.

The majority of prior work on the temporal development or evolution of software
has been based on individual pieces of software, here generically referred to
as "packages". One notably exception is a recent analysis of the "pypi" (Python
Package Index) package manager for python [@pypi2019], which analysed and
presented a number of statistics relevant to the present work, including:

- Compound annual growth rates in numbers of new packages were slightly less
  (at 43%) than growth in active packages (47%) and in new releases (51%).
- Compound annual growth rates in numbers of new package authors was lower than
  these, at 39%.
- Each package has around 7 +/- 19 new releases per year, at an average
  inter-release frequency of 65 +/ 144 days.
- Each author has 2 +/- 17 packages, and 15 +/ 125 releases.
- Median package sizes are 20-40 KB, while mean values are 2-20 MB.
- Total numbers of imported packages increases with compound annual growth rate
  of 62%.
- The standard python libraries represent around 19% of all package imports.

This analyses represents the most comprehensive analysis to date of a package
manager for any computer language, drawn almost two million releases of 179,000
packages. The size of pypi necessarily restricted these analyses to package
metadata, although single metrics of package size were also included. Moreover,
the size of pypi meant that analyses of dependency networks were also
relatively restricted to aggregate statistics only, with no analyses of the
internal properties of dependency networks.

More detailed considerations of temporal development and evolution of code have
largely been restricted to individual packages. Analyses of code evolution
frequently refer to work of @software-evolution-97, who expressed their
findings as a series of "laws" of software evolution. These "laws" include
that, over time, software increases both in complexity and in functional
content (or numbers of functions). Those laws of growth themselves derive from
@Turski96, who posited an inverse-square law for software growth, later
formalised in @Turski02.

These "laws" of software evolution have been frequently used and cited (see
overview in @Mens2008, and references therein), although overwhelmingly for
considerations of individual software packages. They have been widely adopted
in a burgeoning field of what might be referred to as "Software Mining and
Analytics", to elucidate useful predictors of software
quality. A useful overview of the
field was
provided by @mining-repos-2013, who usefully identified a number of categories,
including that of "code statistics" examined here.

More detailed considerations of package dependency networks have been developed
in more restricted, and thus more manageable, contexts, such as the analysis of
the Debian network of @macro-evolution-2009, which revealed progressive
increases in package inter-dependencies. Although they observe increases in the
sizes of packages over time, and claim that this reflects the laws of
@software-evolution-97, they neither quantify nor model actual rates of change.  <!-- do we or do we plan to? -->
They also observe that the relative sizes of packages, measured in terms of
lines of code, remains relatively stable across releases.

A more extensive consideration of dependency networks alone was provided by 
@seven-pkg-networks-2019, who examined networks from seven package management
systems, including pypi and CRAN. They observe
that both numbers of packages and numbers of dependencies increase more slowly
for CRAN packages than for any of the other systems they consider, as do
numbers of package updates. They also observe that the "survival probability"
(that a package is not updated at some time, $t$) <!-- This seems to be reversed? --> is far greater for CRAN than
for other package ecosystems (although with differences observed between
packages which are not required by other packages, and those which are).
Finally, they observe that the Gini coefficient of inequality between dependent
packages has also increased over time in all systems.

This work focusses on the CRAN package system which is around one tenth the
size of pypi for the python language, although still amounting to over 100,000
releases of around 20,000 packages. This smaller sizes enables far more
detailed analyses than previous work, not just of package metadata, but also
extending to detailed analyses of the code itself.

First, we describe the `{pkgstats}` R package, which can be used to generate
and analyze properties of R package code, and the accompanying database of code
statistics of all historical CRAN packages.  We then describe trends in CRAN
R package properties over the history of the archive (199X-2021).  We focus on
trends in four areas: Package metadata (licenses, authorship, translation),
Package and content (file and function types and sizes), intra-package code
structure (function call networks), and finally, dependency networks amongst
packages. Finally, we discuss implications and likely drivers of these trends,
and potential further applications and extensions of the `pkgstats` package
and database.

# Methods

## The `pkgstats` package

<!-- This secion should be way longer. It shoule
- described pkgstats package in more detail, to the point of basic usage
- describe the different metrics collected - if not a full list, then at least
  categories and a reference to a full list
- describe the usage of `ctags` and generation of relevant network statistics
- describe the availability of the full package database and where to download
  it
-->

Most of the methodology used in this study derives from [the `pkgstats`
package](https://docs.ropensci.org/pkgstats/), an R package with
a primary eponymous function used to collect static metrics from an R package.
This package also contains [a function,
`pkgstats_from_archive`](https://docs.ropensci.org/pkgstats/reference/pkgstats_from_archive.html),
which can be applied to a local mirror of the CRAN archive to collate
statistics from all packages. Data for these analyses were derived directly
from this function.

<!-- Then, a brief description of analytical methods, referencing exact variables -->

To describe the evolution of development practices for CRAN packages over time,
we analyze a subset of these metrics. We examine authorship (`field_name`), ....
We calculate changes in amount and distribution of code amongst the different
functional directories in R packages (`R`, `src`, etc.), ....

## Inverse-Square Law of Software Evolution

<!-- I don't think it makes sense to put this in the methods section unless you plan to actually do some goodness-of-fit analysis. -->

One of our aims was to examine the applicability of the inverse-square law of
software evolution [@Turski96; @software-evolution-97; @Turski02]. This law
describes increases in numbers of software "modules" which translate in the
context of R packages to functions. Increases over time in these numbers, $s$,
are posited to follow the inverse square law,

\begin{equation}
    s_i = s_{i - 1} + \bar{E} / s_{i - 1} ^ 2,
\end{equation}
where $\bar{E}$ is a model parameter calculated from,
\begin{equation}
    E_i = (s_i - s_1) / (\sum_{k=1}^{i-1} 1 / s_k^2).
\end{equation}

## Network Analyses

- [ ] Describe / quantify effect of GitHub founding in 2008, as many results
  manifest distinct changes from around 2010 onwards.

# Results

## Metadata

The relatively regulated process of CRAN submissions [@seven-pkg-networks-2019]
is reflected in more regulated meta-data able to be extracted from package
'DESCRIPTION' files. Numbers of both authors and contributors have increased
over time (Fig. 2A), with contributors having been rarely acknowledged at all
prior to 2010 (two years after the launch of GitHub), increasing to an average
of one acknowledged contributor per package in 2021. Numbers of unique licenses
manifest an opposite tendency, decreasing notably
over time (Fig. 2B, resulting in the values shown in Table 1.

```{r fig2, child = "figures/fig2.Rmd", eval = TRUE, echo = FALSE}
```

```{r licenses, echo = FALSE}
datafile <- file.path (here, "data-raw", "pkgstats-results.Rds")
x <- load_pkgstats_data (datafile, raw = TRUE, latest = FALSE)
index <- which (x$year > 2010)
lic <- sort (table (x$license [index]), decreasing = TRUE)
lic <- data.frame (license = names (lic),
                   n = as.integer (lic)) [1:10, ]
knitr::kable (lic, caption = "**Table 1.**")
```

```{r fig3, child = "figures/fig3.Rmd", eval = TRUE, echo = FALSE}
```


Numbers of translations peaked broadly in the 2010's, and have declined notably
since around 2015, suggesting a pronounced increased acceptance since that time
of English as a globally standard language, and a concomitant decrease in
active acknowledgement of other languages (Fig. 3A). Finally, analyses of URLs
provided in package 'DESCRIPTION' files clearly reveals an abrupt break after
around 2010 with the advent of online code repositories. GitHub has always
dominated, while r-forge has declined progressively since around 2015. The most
notable increase has been the emergence of rOpenSci as provider of the only URL
to come close to within 1/20th of the popularity of GitHub over the past few
years.


## Package Structure and Content

### File Structure

The temporal development of numbers of files manifest several notable
characteristics (Figure 4). Unsurprisingly, R packages have always been
dominated by files in the `R/` directory, numbers of which increased considerably
up to around 2010, after which they are generally stabilised at a mean value of
around 20 files (Fig. 4A). The `src` and `data` directories are the only places
within an R package in which numbers of files have remained relatively constant
over the preceding 20+ years. Packages generally had very few files in the
`inst` directory prior to 2010, after which they have notably increased. The
two sub-directories in which numbers of files have increased most significantly
are `vignettes` and `tests`, clearly indicating that packages have become both
better documented and better tested, especially since 2010.


```{r fig4, child = "figures/fig4.Rmd", eval = TRUE, echo = FALSE, warning = FALSE}
```

Figure 4B shows equivalent relative changes within individual packages across
successive releases. The changes there are proportional changes in numbers of
files in each of the nominated directories, starting with the change between
the second and first release (a "Release sequence" of 2). Although numbers of
files generally increase with each new release, *rates* of increase tend to
slow down, with packages approaching stability (a ratio of 1) after around 20
releases.  <!-- if this is the case, you should plot the trend out to 20 releases -->

### Function Number, Type and Size

Figure 7 shows the temporal evolution of a number of code metrics. These figures distinguish between the two primary kinds of R functions, exported and non-exported, and also between the main sub-directories of R packages:

- The `R` directory holding all source files in the `R` language;
- The `src` directory holding all code in other languages which needs to be compiled on package installation;
- The `inst` directory generally holding bundles of external code, but also
  including code linked to within `src` code and used for cross-linking in
  compilation of other packages.

Lines of Code ("LoC") in the primary `R` and `src` directories broadly
increased up to around 2010 or slightly later, after which time LoC in these
directories has progressively decreased. These follow similar trends in LoC per
function (Fig. 7B), excepting the early years of CRAN, during which functions
often had well over 100 lines each, especially `src` functions defined in other
languages.

Total numbers of R functions initially increased until around 2007, peaking at
just under 50 exported and 75-80 non-exported functions per package (Fig. 7C).
Numbers of exported R functions have decreased since then to around 27
functions per package in 2021. In contrast, numbers of functions defined in
`src` directories have progressively increased, and now total just over 150
functions. In both `R` and `src` directories, numbers of functions per source
file have nevertheless progressively decreased since the early 2000's to now
amount to less than six `R` functions and just over 8 `src` functions per file
(Fig. 7D).



@static-analysis-2016,

## Intra-package function call networks


## Dependency Networks

Like the "standard" libraries for the python language, R has a core set of
"base" and "recommended" packages. Figure 5A shows the change in the relative
use of these compared with "contributed", or externally user-generated
packages, over time. The use of recommended packages has remained relatively
constant over time, representing around 15-20% of all function calls. Calls to
base R functions represented over 80% of all function calls in the early 2000s,
yet have declined in recent years to below 70% (for unique function calls).
This reduction has been offset by a concomitantly pronounced increase in calls
to functions from contributed packages.


```{r fig5, child = "figures/fig5.Rmd", eval = TRUE, echo = FALSE}
```

Figure 5B shows the "coupling instability" of packages over time. This is the
proportion of *afferent* to total couplings, where afferent couplings are
numbers of (unique) calls that a package makes to dependent functions defined
in other packages. Total couplings include *efferent* couplings, or calls from
other packages to functions of a given package. This measure of instability is
thus a measure of the relative dependence of a package on external
functionality compared with the observe dependence of other packages on that
package. Packages which depend on many external functions, yet on which no
other package depends are thus "unstable", whereas packages with little or not
dependence on external packages, yet upon which many other packages depend, are
comparably "stable".

This result reveals that instability remained broadly stable until around 2012,
after which it has progressively increased, from around 50% to almost 80%. This
latter figure means that packages make on average eight calls to other packages
for every two calls made to that package.

Figure 5C depicts changes in numbers of unique function calls from each package
to the top ten contributed packages, defined as the ten packages with the
largest total number of unique function calls from all packages throughout the
history of CRAN. Numbers of function calls are shown on a logarithmic scale,
with numbers of calls to each of these packages increasing exponentially since
around 2013. The top two packages of {ggplot2} and {dplyr} account for
`r prop_top2`% of all function calls from all contributed packages over the
time shown.

Figure 6 shows results equivalent to Figs 5A-B, but for changes with
progressive releases of each package. The first panel (Fig. 5A) shows changes
in the proportion of calls to the base and recommended packages for each
sequential release, averaged over all packages. (Equivalent proportions of
calls to recommended packages stay broadly constant, albeit with marked
variability, and are not shown.) Relative changes generally remain above 1,
indicating that packages make progressively more calls to base and recommended
packages as they mature, with rates of calls to recommended packages
progressively decreasing, and approaching stability after around 30 releases.

Calls to the base R package initially increase with each release out to around
10-15 releases, indicating that packages incorporate progressively more calls
to base R functions with each release. Beyond around 15 releases, relative
rates of change also approach the stable level of 1, as for calls to
recommended packages, with no further calls being added.

Figure 6B shows the average trajectory of coupling instability for individual
packages across the depicted numbers of sequential releases. While decreases in
coupling instability should be expected as packages mature and are imported by
increasing numbers of other packages (that is, afferent coupling should
generally increase), that figure also shows that mean numbers of efferent calls
tend to decrease with each sequential release. Numbers of afferent calls (that
is, inward calls from other packages to a given package) also naturally
increase with sequential releases, as increasing numbers of packages depend on
a given package. The observed decrease in instability reflects this combination
both of increasing dependence of other packages on any given package, and
decreasing dependence of packages on functionality provided by external
packages.

Finally, Fig. 6C shows progressions of packages being imported into other
packages as a function of release sequence. The number of packages importing
a package increases progressively with numbers of releases. Packages with less
than ten releases are, on average, imported by very few other packages, while
packages with 30 or more releases are imported by ten other packages on
average. The second line on that figure nevertheless reveals that numbers of
functions actually used by other packages increase above one after around ten
releases, yet always remain below two on average, even for packages with 50 or
more releases.

```{r fig6, child = "figures/fig6.Rmd", eval = TRUE, echo = FALSE, warning = FALSE}
```

## Temporal Development #3: Code Metrics

Figure 7 shows the temporal evolution of a number of code metrics. These figures distinguish between the two primary kinds of R functions, exported and non-exported, and also between the main sub-directories of R packages:

- The `R` directory holding all source files in the `R` language;
- The `src` directory holding all code in other languages which needs to be compiled on package installation;
- The `inst` directory generally holding bundles of external code, but also
  including code linked to within `src` code and used for cross-linking in
  compilation of other packages.

### Annual Developments

Lines of Code ("LoC") in the primary `R` and `src` directories broadly
increased up to around 2010 or slightly later, after which time LoC in these
directories has progressively decreased. These follow similar trends in LoC per
function (Fig. 7B), excepting the early years of CRAN, during which functions
often had well over 100 lines each, especially `src` functions defined in other
languages.

Total numbers of R functions initially increased until around 2007, peaking at
just under 50 exported and 75-80 non-exported functions per package (Fig. 7C).
Numbers of exported R functions have decreased since then to around 27
functions per package in 2021. In contrast, numbers of functions defined in
`src` directories have progressively increased, and now total just over 150
functions. In both `R` and `src` directories, numbers of functions per source
file have nevertheless progressively decreased since the early 2000's to now
amount to less than six `R` functions and just over 8 `src` functions per file
(Fig. 7D).

```{r fig7, child = "figures/fig7.Rmd", eval = TRUE, echo = FALSE, warning = FALSE}
```

## Intra-package Developments

Figure 8 shows results equivalent to Fig. 7, but for the development of the
respective metrics with each sequential release of a package, rather than being
averaged over all packages for each year. Lines of code generally increase as
packages mature, with mean numbers of lines in the `R` directory increasing
tenfold from under 1000 for first releases to almost 10,000 in packages with
100 or more releases. Changes in LoC in `src` and `inst` directories were less
clear, yet the dominance of `R` code nevertheless produced an overall increase
in LoC out to the 100th release of packages of around 12 times.

These increases in LoC nevertheless corresponded to an equivalent decrease in
LoC per function with increasing numbers of releases (Fig. 8B). Similar to the
aggregate LoC measures of panel A, LoC per function in `src` directories
remained broadly stable over releases, while LoC per `R` function decreased
notably, and more so for exported than non-exported functions. Exported
functions after 100 releases had just over 30 LoC per function, whereas initial
releases average just under 60 LoC.

The increasing total measures of LoC of panel A combined with decreasing LoC
per function of panel B must mean that more functions are added as packages
mature, which is precisely what Fig. 8C reveals. Packages initially contain
around 20 exported functions on average, yet this increases almost 13-fold to
over 250 by the 100th release. Non-exported functions increase by a similar
10-fold proportion over the first 100 releases, while packages start with more
many more `src` functions on initial releases, and so increases relatively less
as packages mature. Finally, numbers of functions per file show no clear trend
for `src` code, while approximately doubling in `R` code over the first 100
releases of a package, from around 7 to 15.

```{r fig8, child = "figures/fig8.Rmd", eval = TRUE, echo = FALSE, warning = FALSE}
```

## Network metrics

```{r fig9, child = "figures/fig9.Rmd", eval = TRUE, echo = FALSE, warning = FALSE}
```


# References
