---
title: "pkgstats"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        number_sections: true
        theme: flatly
bibliography: pkgstats.bib
header-includes: 
    - \usepackage{tikz}
    - \usetikzlibrary{arrows}
vignette: >
  %\VignetteIndexEntry{pkgstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r pkg-load, echo = FALSE, message = FALSE}
library (pkgstatsAnalyses)
library (ggplot2)
library (patchwork)
library (dplyr)
library (tidyr)
library (igraph)
here <- here::here ()
```

```{r dl-data, echo = FALSE}
path <- file.path (here, "data-raw")
if (!dir.exists (path)) {
    dir.create (path, recursive = TRUE)
}

u <- paste0 ("https://github.com/mpadge/pkgstats-analyses/",
             "releases/download/v0.0.1/")

f_r <- file.path (path, "pkgstats-results.Rds")
if (!file.exists (f_r)) {
    u_r <- paste0 (u, "pkgstats-results.Rds")
    download.file (u_r, f_r)
}

f_p <- file.path (path, "pypi.Rds")
if (!file.exists (f_p)) {
    u_r <- paste0 (u, "pypi.Rds")
    download.file (u_p, f_p)
}
```


# Introduction

A key component of the success of Open Source Software (OSS) is package
managers [@pkg-mngrs-2018], which effectively serve as curated repositories of
independently-contributed software. Package managers exist both for specific
operating systems [@pkg-mngrs-2018] and specific computer languages such as
python [@new-pypi-2018], LaTeX [@ctan-wiki], JavaScript [@npm-2020], and many
others. This work analyses all current and historically archived packages in
the CRAN (Comprehensive R Archive Network) package manager for the R language.

The majority of prior work on the temporal development or evolution of software
has been based on individual pieces of software, here generically referred to
as "packages". One notably exception is a recent analysis of the "pypi" (Python
Package Index) package manager for python [@pypi2019], which analysed and
presented a number of statistics relevant to the present work, including:

- Compound annual growth rates in numbers of new packages were slightly less
  (at 43%) than growth in active packages (47%) and in new releases (51%).
- Compound annual growth rates in numbers of new package authors was lower than
  these, at 39%.
- Each package has around 7 +/- 19 new releases per year, at an average
  inter-release frequency of 65 +/ 144 days.
- Each author has 2 +/- 17 packages, and 15 +/ 125 releases.
- Median package sizes are 20-40 KB, while mean values are 2-20 MB.
- Total numbers of imported packages increases with compound annual growth rate
  of 62%.
- The standard python libraries represent around 19% of all package imports.

This analyses represents the most comprehensive analysis to date of a package
manager for any computer language, drawn almost two million releases of 179,000
packages. The size of pypi necessarily restricted these analyses to package
metadata, although single metrics of package size were also included. Moreover,
the size of pypi meant that analyses of dependency networks were also
relatively restricted to aggregate statistics only, with no analyses of the
internal properties of dependency networks.

More detailed considerations of temporal development and evolution of code have
largely been restricted to individual packages. Analyses of code evolution
frequently refer to work of @software-evolution-97, who expressed their
findings as a series of "laws" of software evolution. These "laws" include
that, over time, software increases both in complexity, and in functional
content (or numbers of functions). Those laws of growth themselves derive from
@Turski96, who posited an inverse-square law for software growth, later
formalised in @Turski02.

These "laws" of software evolution have been frequently used and cited (see
overview in @Mens2008, and references therein), although overwhelmingly for
considerations of individual software packages. They have been widely adopted
in a burgeoning field of what might be referred to as "Software Mining and
Analytics", albeit very commonly to elucidate useful predictors of software
quality, rather than on general patterns of evolution. A useful overview of the
field, in the form of an extensive meta-review of extant literature, was
provided by @mining-repos-2013, who usefully identified a number of categories,
including that of "code statistics" examined here.

More detailed considerations of package dependency networks have been developed
in more restricted, and thus more manageable, contexts, such as the analysis of
the Debian network of @macro-evolution-2009, which revealed progressive
increases in package inter-dependencies. Although they observe increases in the
sizes of packages over time, and claim that this reflects the laws of
@software-evolution-97, they neither quantify nor model actual rates of change.
They also observe that the relative sizes of packages, measured in terms of
lines of code, remains relatively stable across releases.

A more extensive consideration of dependency networks alone was provided by 
@seven-pkg-networks-2019, who examined networks from seven package management
systems, including pypi and CRAN, the focus of the present work. They observe
that both numbers of packages and numbers of dependencies increase more slowly
for CRAN packages than for any of the other systems they consider, as do
numbers of package updates. They also observe that the "survival probability"
(that a package is not updated at some time, $t$) is far greater for CRAN than
for other package ecosystems (although with differences observed between
packages which are not required by other packages, and those which are).
Finally, they observe that the Gini coefficient of inequality between dependent
packages has also increased over time in all systems.

This work focusses on the CRAN package system which is around one tenth the
size of pypi for the python language, although still amounting to over 100,000
releases of around 20,000 packages. This smaller sizes enables far more
detailed analyses than previous work, not just of package metadata, but also
extending to detailed analyses of the code itself.

<!--
# Research Questions
-->

# Methods

Most of the methodology used in this study derives from [the `pkgstats`
package](https://docs.ropensci.org/pkgstats/), an R package with
a primary eponymous function used to collect static metrics from an R package.
This package also contains [a function,
`pkgstats_from_archive`](https://docs.ropensci.org/pkgstats/reference/pkgstats_from_archive.html),
which can be applied to a local mirror of the CRAN archive to collate
statistics from all packages. Data for these analyses were derived directly
from this function.

## Impact of Covid19 Pandemic on Software Evolution

The analyses here focus on the temporal development of software evolution. Like
almost all aspects of human endeavour, activities of software evolution were
greatly impacted by the advent of Covid19. We begin our analyses with a brief
examination of this temporal disruption to developmental trajectories, doing so
through comparison with equivalent trajectories for python development. These
comparisons are made on the basis of package submissions only, with python data
collected directly from the pypi archive, through obtaining the list of all
pypi packages from [pypi.org/simple](https://pypi.org/simple/), then extracting
json-formatted metadata for each package from the pypi API.

## Inverse-Square Law of Software Evolution

One of our aims was to examine the applicability of the inverse-square law of
software evolution [@Turski96; @software-evolution-97; @Turski02]. This law
describes increases in numbers of software "modules" which translate in the
context of R packages to functions. Increases over time in these numbers, $s$,
are posited to follow the inverse square law,

\begin{equation}
    s_i = s_{i - 1} + \bar{E} / s_{i - 1} ^ 2,
\end{equation}
where $\bar{E}$ is a model parameter calculated from,
\begin{equation}
    E_i = (s_i - s_1) / (\sum_{k=1}^{i-1} 1 / s_k^2).
\end{equation}

## Network Analyses

- [ ] Describe / quantify effect of GitHub founding in 2008, as many results
  manifest distinct changes from around 2010 onwards.

# Results

## Impact of Covid19 Pandemic on Software Evolution

Submission rates to pypi were largely unaffected by Covid19, in stark contrast
to equivalent submission rates to CRAN (Fig. 1A).

```{r pre-process-pypi, echo = FALSE, eval = !file.exists ("pypi-prepro.Rds")}
datafile <- file.path (here, "data-raw", "pypi.Rds")
x <- load_pkgstats_data (datafile, raw = FALSE, latest = FALSE)
saveRDS (x, "pypi-prepro.Rds")
```
```{r fig1, child = "figures/fig1.Rmd", eval = TRUE, echo = FALSE}
```


Further examination of the CRAN data to distinguish between new submissions
versus updates of existing packages reveals that the marked downturn evident in
Fig. 1A was due to a decrease in the rate of package updates, while rates of
submission of new packages remained fairly constant (Fig. 1B).

We do not intend to hypothesise any reason for these observed patterns, but
merely note them here for reference in the context of subsequent results. For
reasons we do not speculate on here, updates to R packages have been strongly
impacted by Covid19, unlike equivalent rates either for new submissions to
CRAN, or for rates of submission of python packages to pypi. 

## Temporal Development #1: Packages and meta-data

The relatively regulated process of CRAN submissions [@seven-pkg-networks-2019]
is reflected in more regulated meta-data able to be extracted from package
'DESCRIPTION' files. Numbers of both authors and contributors have increased
over time (Fig. 2A), with contributors having been rarely acknowledged at all
prior to 2010 (two years after the launch of GitHub), increasing to an average
of one acknowledged contributor per package in 2021. Numbers of unique licenses
manifest an opposite tendency, decreasing notably
over time (Fig. 2B, resulting in the values shown in Table 1.

```{r fig2, child = "figures/fig2.Rmd", eval = TRUE, echo = FALSE}
```

```{r licenses, echo = FALSE}
datafile <- file.path (here, "data-raw", "pkgstats-results.Rds")
x <- load_pkgstats_data (datafile, raw = TRUE, latest = FALSE)
index <- which (x$year > 2010)
lic <- sort (table (x$license [index]), decreasing = TRUE)
lic <- data.frame (license = names (lic),
                   n = as.integer (lic)) [1:10, ]
knitr::kable (lic, caption = "**Table 1.**")
```

```{r fig3, child = "figures/fig3.Rmd", eval = TRUE, echo = FALSE}
```


Numbers of translations peaked broadly in the 2010's, and have declined notably
since around 2015, suggesting a pronounced increased acceptance since that time
of English as a globally standard language, and a concomitant decrease in
active acknowledgement of other languages (Fig. 3A). Finally, analyses of URLs
provided in package 'DESCRIPTION' files clearly reveals an abrupt break after
around 2010 with the advent of online code repositories. GitHub has always
dominated, while r-forge has declined progressively since around 2015. The most
notable increase has been the emergence of rOpenSci as provider of the only URL
to come close to within 1/20th of the popularity of GitHub over the past few
years.


### Internal Package and File Structure

The temporal development of numbers of files manifest several notable
characteristics (Figure 4). Unsurprisingly, R packages have always been
dominated by files in the R directory, numbers of which increased considerably
up to around 2010, after which they are generally stabilised at a mean value of
around 20 files (Fig. 4A). The `src` and `data` directories are the only places
within an R package in which numbers of files have remained relatively constant
over the preceding 20+ years. Packages generally had very few files in the
`inst` directory prior to 2010, after which they have notably increased. The
two sub-directories in which numbers of files have increased most significantly
are `vignettes` and `tests`, clearly indicating that packages have become both
better documented and better tested, especially since 2010.


```{r fig4, child = "figures/fig4.Rmd", eval = TRUE, echo = FALSE, warning = FALSE}
```

Figure 4B shows equivalent relative changes within individual packages across
successive releases. The changes there are proportional changes in numbers of
files in each of the nominated directories, starting with the change between
the second and first release (a "Release sequence" of 2). Although numbers of
files generally increase with each new release, *rates* of increase tend to
slow down, with packages approaching stability (a ratio of 1) after around 20
releases.


@static-analysis-2016,

## Temporal Development #2: Dependency Networks

Like the "standard" libraries for the python language, R has a core set of
"base" and "recommended" packages. Figure 5A shows the change in the relative
use of these compared with "contributed", or externally user-generated
packages, over time. The use of recommended packages has remained relatively
constant over time, representing around 15-20% of all function calls. Calls to
base R functions represented over 80% of all function calls in the early 2000s,
yet have declined in recent years to below 70% (for unique function calls).
This reduction has been offset by a concomitantly pronounced increase in calls
to functions from contributed packages.


```{r fig5, child = "figures/fig5.Rmd", eval = TRUE, echo = FALSE}
```

Figure 5B shows the "coupling instability" of packages over time. This is the
proportion of *afferent* to total couplings, where afferent couplings are
numbers of (unique) calls that a package makes to dependent functions defined
in other packages. Total couplings include *efferent* couplings, or calls from
other packages to functions of a given package. This measure of instability is
thus a measure of the relative dependence of a package on external
functionality compared with the observe dependence of other packages on that
package. Packages which depend on many external functions, yet on which no
other package depends are thus "unstable", whereas packages with little or not
dependence on external packages, yet upon which many other packages depend, are
comparably "stable".

This result reveals that instability remained broadly stable until around 2012,
after which it has progressively increased, from around 50% to almost 80%. This
latter figure means that packages make on average eight calls to other packages
for every two calls made to that package.

Figure 5C depicts changes in numbers of unique function calls from each package
to the top ten contributed packages, defined as the ten packages with the
largest total number of unique function calls from all packages throughout the
history of CRAN. Numbers of function calls are shown on a logarithmic scale,
with numbers of calls to each of these packages increasing exponentially since
around 2013. The top two packages of {ggplot2} and {dplyr} account for
`r prop_top2`% of all function calls from all contributed packages over the
time shown.

- Expansion of base + recommended packages over time
- Use of base + recommended package in other packages over time
- Use of contributed packages over time

(These can not be done until data re-generated following [`pkgstats`
issue#32](https://github.com/ropensci-review-tools/pkgstats/issues/32).

## Temporal Development #3: Code Metrics

- Computer Languages
- Trends in code metrics
- Inter- versus intra-package patterns

# References
